A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-01 paper.tex
Intro draft

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. [THIS IS BAD, WHY?]
We also lack a shared format for interchanging ARG data [which is bad
for users, progress]. [Because we don't really agree on what an ARG is,
we don't really know what properties it should have, when we infer it].

% What do we do?
Here we precisely define an encoding for ARGs based on specific genomes
and their intervals of genetic inheritance. We contrast this ``gARG''
encoding with the classical approach derived from the coalescent
with recombination stochastic process of Griffiths
and colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}. As this is based
on recording evolutionary events, we refer to it as the event ARG, or
``eARG'' encoding.
We show that the gARG encoding is both flexible and efficient,
and can represent varying levels of precision about historical
events.
% Wilder: important to get this point into the intro
We argue that while adopting the general concept of an ARG data structure
is useful, it is important that we then fully decouple this data structure
from the original stochastic process.
[MORE]

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the existance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is not as simple as it may seem.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.

>>>>

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

A general term to describe the detailed genetic genealogy of a sample 
that encompasses the output of all these methods is indeed useful, but 
there are substantial problems emerging from the current usage of the term 
ARG (and its implicit concrete definition in the Griffiths formulation).
The problem is much deeper than some pedantic discussion over terminology,
and directly informs what we should be attempting to estimate, how we 
evaluate inference methods, and what the limitations on the precision
and accuracy of our inferences are.
We currently lack a well-defined target for inference, and therefore there is 
disagreement on how these inference should be evaluated, and 
what the criteria for success should be.
We also lack a shared format for interchanging ARG data between
programs, with the assorted negative consequences which
geneticists are only too familiar with~\citep{excoffier2006computer}.
Finally, and perhaps most importantly, and overly narrow definition of ARGs
leads to potentially missed opportunities in their application and usage.

% What do we do?
In this paper we attempt to address these problems by 
providing a precise definition of an ARG that is general enough to encompass
the output of all currently available inference methods. 
The detailed genetic genealogy of a sample is defined in terms 
of based on specific genomes
and their intervals of genetic inheritance, which we refer to as 
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach which 
is based on recording evolutionary events, and we therefore refer to as 
the ``event ARG'', or ``eARG'' encoding.
We show that the gARG encoding has a range of advantages over the 
eARG approach, being both flexible and efficient.
In particular, we argue that the goal of inferring fully precise 
ARGs---a necessity under the eARG encoding---should be critically 
examined as part of the process of fully decoupling our ARG data 
structure from the coalescent with recombination model.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the existance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
